---
title: "An introduction to mrmr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro-to-mrmr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width = 7, 
  fig.height = 3.5
)
```

The `mrmr` package is designed to simplify and expedite mark recapture data 
processing, parameter estimation, and visualization for a narrow class of 
mark recapture data (namely, those collected by the Sierra Nevada Amphibian 
Research Group).

# Workflow overview

The typical workflow involves three steps:

1. Cleaning mark recapture data with `mrmr::clean_data()`
2. Fitting a mark recapture model with `mrmr::fit_model()`
3. Visualizing results with `mrmr::plot_model()`

## Data specifications

The package comes with example data that illustrate the expected format. 
Three files are needed. 

The first file specifies capture data:

```{r setup}
library(dplyr)
library(readr)
library(tibble)
library(rstan)
library(mrmr)
```

```{r show-capture-data}
captures <- system.file("extdata", 
                                "capture-example.csv", 
                                package = "mrmr") %>%
  read_csv

glimpse(captures)
```

Note that the `system.file` function generates a path to the 
`capture-example.csv` file, but in a real application, you would likely use a 
path to a local file (real data, not the example data included in the package).
The files must be formatted with the same columns and conventions uesd in the 
example data. 
Each row in this file represents a capture or recapture event.

The second file specifies translocation data: 

```{r show-trans-data}
translocations <- system.file("extdata", 
                                      "translocation-example.csv", 
                                      package = "mrmr") %>%
  read_csv

glimpse(translocations)
```

Each row corresponds to a translocation event of one unique individual. 

The third file specifies survey-level data:

```{r show-survey-data}
surveys <- system.file("extdata", 
                               "survey-example.csv", 
                                package = "mrmr") %>%
  read_csv

glimpse(surveys)
```

Here, each row is a unique survey.

## Data processing

To load and process the data, provide paths to the data files as 
arguments to the `mrmr::clean_data()` function:

```{r clean-data, message=FALSE}
data <- clean_data(captures, surveys, translocations)
```

The output provides data frames as list elements, and a list element called
`stan_d`, which contains data that has been pre-processed for model fitting. 

### Including detection/non-detection covariates

Survey-level covariates that might affect the probability of detection can 
be included via the `capture_formula` argument to `clean_data`. 
**Note:** it is a good idea to ensure that any continuous covariates are 
rescaled prior to using them as covariates to have mean = 0 and unit standard 
deviation. 
This can be acheived by adding a column to the `captures` data
Using a covariate value like `year`, which typically has a mean in the thousands
and a standard deviation far from one will almost assurredly result in 
numerical issues when fitting a model. 

Assuming we wanted to include a covariate for detection like `person_hours`, 
we can do so as follows:

```{r detection-covariate, message=FALSE, eval=FALSE}
# center and scale the covariate
captures <- captures %>%
  mutate(c_person_hours = c(scale(person_hours)))

# specify the formula
data <- clean_data(captures, surveys, translocations,
                   capture_formula = ~ c_person_hours)
```

### Including survival covariates

Individual-level survival covariates can be included optionally via the 
`survival_formula` and `survival_fill_value` arguments. 
Both of these must be specified, because covariate values for pseudo-individuals 
must be filled in (they are never observed).
So, for example, to evaluate the effect of an experimental treatment, if some 
individuals belong to a "treatment" group and others belong to an "control" 
group, then a group must be specified as a fill value (e.g., "wild-caught" or 
"control", depending on the experiment):

```{r survival-covariate, message=FALSE, eval=FALSE}
# specify the formula
data <- clean_data(captures, surveys, translocations,
                   survival_formula = ~ treatment, 
                   survival_fill_value = c(treatment = "wild-caught"))
```


## Model structure

The `mrmr` package implements a Bayesian open-population Jolly-Seber mark recapture model with known additions to the population (introduced adults).
The model tracks the states of $M$ individuals that comprise a superpopulation made up of real and pseudo-individuals (see Joseph and Knapp 2018, *Ecosphere* for details).

We assume a robust sampling design where the states of individuals are constant within primary periods, which have repeat secondary periods within them (during which observations are made). 
The possible states of individuals include "not recruited", "alive", and "dead". 
The possible observations of individuals include "detected" and "not detected". 
We assume that individuals that are in the "not recruited" or "dead" states are never detected (i.e., there are no mistakes in the individual PIT tag ID records).

## Model fitting

To fit a mark-recapture model, use `mrmr::fit_model()`. 
This model accounts for known introductions into the population, and has 
random effects to account for variation in survival and recruitment through time. 
The example below uses two chains and relatively few iterations for speed, but 
it is a good idea to run 3+ chains, and at least 1000 iterations.

```{r fit-model}
mod <- fit_model(data, cores = parallel::detectCores(), 
                 chains = 2, iter = 500)
```

To diagnose potential non-convergence of the MCMC algorithm, inspect traceplots:

```{r plot-traceplots}
pars_to_plot <- c('alpha_lambda', 
                  'sigma_lambda', 
                  'beta_phi', 
                  'sigma_phi', 
                  'beta_detect')
traceplot(mod$m_fit, pars = pars_to_plot)
```

It is also a good idea to look at Rhat estimates to see whether they 
indicate of a lack of convergence (Rhat values $\geq$ 1.01):

```{r print-model}
print(mod$m_fit, pars = pars_to_plot)
```

## Built-in visualizations

Time series of abundance, recruitment, and survival of introduced cohorts are
available through the `mrmr::plot_model()` function. 

```{r plot-abundance}
plot_model(mod, what = 'abundance')
```

```{r plot-recruitment}
plot_model(mod, what = 'recruitment')
```

```{r plot-survival}
plot_model(mod, what = 'survival')
```

## Custom visualizations

Any of the plotting functionality that you would expect from a `stanfit` model
is available as well, by accessing the `m_fit` list element from a model object. 
For example, we could assess the posterior for the superpopulation size:

```{r}
traceplot(mod$m_fit, pars = "Nsuper")
```

## Cases with no translocations

In some cases, there may not be any introductions of animals into a population. 
The only difference in implementation in such cases is that the `translocations` 
argument to `clean_data()` can be omitted:

```{r no-trans, message = FALSE, eval = FALSE}
data_no_trans <- clean_data(captures, surveys)
mod_no_trans <- fit_model(data_no_trans)
```

## Cases with known removals

When individuals are removed from a population, e.g., for translocation, these
known removals can be included as data to increase the precision of abundance
estimates. 
An example dataset fitting this case is included at an imaginary site
called "Equid". 
Note below that we filter out dead animals -- any dead animals encountered on 
surveys should be in the `removals` data, and not in the `captures` data frame, 
which should only contain live animals (those that might survive to the next
primary period).

```{r removals, message = FALSE, eval = FALSE}
captures <- system.file("extdata", 
                        "equid-captures.csv", 
                         package = "mrmr") %>%
  read_csv %>%
  filter(capture_animal_state != 'dead')
surveys <- system.file("extdata", 
                       "equid-surveys.csv", 
                       package = "mrmr") %>%
  read_csv
removals <- system.file("extdata", 
                        "equid-removals.csv", 
                         package = "mrmr") %>%
  read_csv

d <- clean_data(captures, surveys, removals = removals)
m <- fit_model(d, cores = 2, chains = 2)
```